---
title: "VSV 2023 California Outbreak"
description: " "
date: "2025-12-18"
categories: [R, VSNJ, Phylogeography]
output-dir: docs
format:
  gfm:
   toc: true
   toc-depth: 2
   code-fold: show
   code-summary: "Hide code"
   preview-mode: raw
editor: source
editor_options:  
  chunk_output_type: console
---

## Preliminaries
### Libraries
```{r warning=FALSE, message=FALSE, echo=TRUE}
library(here) # directory managment
library(tidyverse) # data wrangling
library(scales) # supports ggplot2

#spatial
library(ggmap) # mapping
library(ggspatial) # mapping
library(sf) # spatial data manipulation
library(tigris) # geographic data
library(terra) # spatial data manipulation
library(rnaturalearth) # global geographic data
library(rnaturalearthhires) # global geographic data
library(rnaturalearthdata) # global geographic data

# Phylo
library(ape) #Analyses of Phylogenetics and Evolution (APE)
library(phangorn) # phylogenetic trees and networks
library(treeio) # manipulate trees
library(Biostrings) # sequence wrangling
library(msa) # Multiple Sequence Alignment (MSA) algorithms  
library(ggtree) # tree visualization and annotation
library(EpiSky) # prevalence and Re estimation
```


### Custom Functions
Load custom functions
```{r}
source(here("R/utilities.R"))
source_dir(here("R"))
```


### Register API
Register Stadia Maps API to pull background images.
```{r}
map_api <- yaml::read_yaml(here("local", "secrets.yaml"))
register_stadiamaps(key = map_api$stadi_api)
```

## Geographic Boundaries

### California Counties
Jurisdictional boundaries for California, USA counties
```{r}
options(tigris_use_cache = TRUE)

ca_counties <- counties(state = "CA", year = 2022, class = "sf")
```

### Get Centroids
Obtain geographic coordinates for county centers
```{r}
ca_centroids_sf <- st_centroid(ca_counties)

# to SpatVector
ca_centroids_sv <- vect(ca_centroids_sf)

coord_frame <- as.data.frame(ca_centroids_sv, geom="xy") %>%
  mutate(county = NAME) %>%
  select(county, x, y)
```

## VSV Data

### Unsequenced Detections
```{r message=FALSE, warning=FALSE}
vsv_cases <- read_csv(here("local/vsv_county_noPII.csv")) %>%
  filter(ONSET_YEAR >= 2023 & ONSET_YEAR <= 2024) %>%
  mutate(date = paste0(ONSET_YEAR, "-", ONSET_MONTH, "-", ONSET_DAY),
         date = as_date(date, format = "%Y-%m-%d"))

us_cases <- vsv_cases %>%
  filter(COUNTRY == "USA") 

mex_states <- c("SON", "CHH", "CHI") # northern states

mx_cases <- vsv_cases %>%
  filter(COUNTRY == "MEX",
         STATE %in% mex_states)

comb_cases <- rbind(us_cases, mx_cases) %>%
  group_by(COUNTRY, date) %>%
  summarise(inc = length(date))


mex_seq_sample <- as_date("2023-01-15") # date of Sonora, Mx sequence
epi_start <- min(comb_cases$date) # inclusive of MX sequence, 2023-01-15
epi_end <- max(comb_cases$date)

epi_phy_lag <- as.integer(mex_seq_sample - epi_start)
epi_phy_lag # days offset between case reports and genomic sample collection
```


## Case Data
Mexico detections are for Sonora and Chihuahua, Mexico only.
```{r fig.width=15, fig.height=10}
inc_plot <- comb_cases %>%
  mutate(
    epiweek_start = floor_date(date, unit = "week", week_start = 1)
  ) %>%
  group_by(COUNTRY, epiweek_start) %>%
  summarise(inc = sum(inc, na.rm = TRUE), .groups = "drop")

ggplot(inc_plot, aes(x = epiweek_start, y = inc, fill = COUNTRY)) +
  geom_bar(stat="identity") +
  scale_fill_manual(
    values = c(
      "USA" = "steelblue",
      "MEX" = "darkorange"
    )
  ) +
  scale_x_date(expand = c(0, 0)) +
  labs(
    fill = "Country",
    x = " ",
    y = "Incidence",
    title = " "
  ) +
  theme_minimal() +
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"),
    legend.position = "bottom",
    legend.text = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 20, face = "bold", vjust = -2),
    axis.title.y = element_text(size = 20, face = "bold", vjust = 3),
    axis.text.x = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 22, face = "bold", colour="gray50", hjust = 0.5)
  )
```

### Genomic Metadata
Metadata associated with genomic sequences
```{r}
vsv_meta <- readxl::read_excel(here("local/Filesnames 1.xlsx"), .name_repair="universal")
names(vsv_meta) <- c("fastq", "meta", "tip", "horse", "date", "county")

vsv_meta <- vsv_meta %>%
  mutate(date = as_date(date),
         tip = gsub("-", "_", tip)) %>%
  select(tip, date, county)

head(vsv_meta)

#get labels
keep_labels <- c(vsv_meta$tip, "Mexico2023") # also keep MX sample

length(keep_labels)
```

### Load and Filter Alignment
Read alignment
```{r}
alignment <- read.dna(here("local/world.fasta"),
                      format="fasta")

# label alignment
matched_labels <- keep_labels[keep_labels %in% rownames(alignment)]
selected_seqs <- alignment[matched_labels, , drop = FALSE]
```

```{r eval=FALSE}
# save copy of alignment
write.nexus.data(as.character(selected_seqs), 
                 file = here("local/vsv_2023.nex"), 
                 format = "dna")
```

### Organize Data
A couple of mispelled county names in the data
```{r}
# correct county names
vsv_meta <- vsv_meta %>%
  mutate(
    county = case_when(
      county == "San Bernadino" ~ "San Bernardino",
      county == "San Luis" ~ "San Luis Obispo",
      tip == "Tulare_10_2" ~ "Tulare",
      TRUE ~ county
    ),
    date = if_else(is.na(date) & tip == "Santa_Barbara", as.Date("2023-07-20"), date)
  )

vsv_meta <- left_join(vsv_meta, coord_frame, by = "county")

# Texas data
vsv_meta <- vsv_meta %>%
  mutate(x = if_else(county == "Shakelford", -99.2786, x),
         y = if_else(county == "Shakelford", 32.8242, y))

mex_meta <- as.data.frame( 
  cbind(
  tip = "Mexico2023",
  date = "2023-01-15",
  county = "Sonora",
  x = -110.3309, # center of Sonora
  y = 29.2972
  )
)

mex_meta$date <- as_date(mex_meta$date)

vsv_meta <- rbind(mex_meta, vsv_meta)

vsv_meta <- vsv_meta %>%
  mutate(x = as.numeric(x),
         y = as.numeric(y))
```

## Substitution Model
Compare models of evolution
```{r}
aligned_phyDat <- as.phyDat(selected_seqs)
  
# run the test, compare the models
mt <- modelTest(aligned_phyDat)

mt %>% # no important differences
  arrange(AIC) %>%
  slice_head(n=5) 

# select model
env <- attr(mt, "env")
best_mod <- eval(get("GTR+I", env), env) 
```

## Maximum Likelihood Tree
Constructing  a quick max likelihood tree to check initial results before inferring time calibrated tree.  
```{r}
fit1 <- optim.pml(best_mod, # best model 
                 optNni = FALSE, optBf = TRUE, 
                 optQ = TRUE, optInv = TRUE, 
                 optGamma = FALSE, optEdge = FALSE, 
                 optRate = TRUE, 
                 control = pml.control(epsilon = 1e-08,
                                       maxit = 200, trace = 0))

#Fix substitution model and fit tree
fit2 <- optim.pml(fit1, 
                 optNni = TRUE, optBf = FALSE,
                 optQ = FALSE, optInv = FALSE, 
                 optGamma = FALSE, optEdge = TRUE,
                 control = pml.control(epsilon = 1e-08, 
                                       maxit = 200, trace = 0))

#Fine tune
fit3 <- optim.pml(fit2, 
                 optNni = TRUE, optBf = TRUE,
                 optQ = TRUE, optInv = TRUE, 
                 optGamma = FALSE, optEdge = TRUE, 
                 optRate = FALSE,
                 control = pml.control(epsilon = 1e-08, 
                                       maxit = 200, trace = 0))
```

## Bootstrap
Bootstrap support values
```{r message=FALSE, warning=FALSE, eval=FALSE}
set.seed(1976)
boots <- bootstrap.pml(fit3,
                       bs = 500, # reduced count for rendering
                       optNni = TRUE,
                       control = pml.control(trace = 0))

saveRDS(boots, here("local/ml_boots_2025-12-15.rds"))
```

### View ML Tree
Node values suggest high uncertainty, likely do to high-relatedness.  
```{r fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
boots <- readRDS(here("local/ml_boots_2025-12-15.rds"))

ml_tree <- fit3$tree

bs_tree <- plotBS(midpoint(ml_tree), boots, 
       type="p", cex=0.4,
       bs.adj = c(1.25, 1.25),
       bs.col = "black")

plotBS(bs_tree)
add.scale.bar()
title("Maximum Likelihood")
```
  
## Time Calibrated Tree
  
### Temporal Signal
Determine if time correlation is present in the sequences (yes)
```{r message=FALSE, warning=FALSE, fig.height=6, fig.width=6}
check_tree <- midpoint(ml_tree)
rtips <- adephylo::distRoot(check_tree)

vsv_meta$date_dec <- decimal_date(vsv_meta$date)

date_vector <- setNames(vsv_meta$date_dec, vsv_meta$tip)
date_vector <- date_vector[check_tree$tip.label]


plot(date_vector, rtips, xlab = "Sampling date", ylab = "Root-to-tip distance")
abline(lm(rtips ~ date_vector), col = "red")
summary(lm(rtips ~ date_vector))
```

Save dates file for BEAST
```{r eval=FALSE}
dates_file <- vsv_meta %>%
  select(tip, date_dec)

write.table(dates_file,
            file = here("local/vsv_dates.txt"),
            sep = "\t",
            row.names = FALSE,
            col.names = TRUE,
            quote = FALSE)
```


### Tracer Stats
Tree run in BEAST
```{r message=FALSE, warning=FALSE}
tracer_stats <- get_tracer_stats(here("local/run_2025-12-18/vsv_2023_geog.log"),
                                 use_burnin=0.10)

stats_to_return <- c("posterior", "likelihood", "prior", "treeLikelihood", 
                     "Tree.height", "Tree.treeLength", "lockRate.vsv_2023",
                     "popSize", "CoalescentConstant")

tracer_stats %>%
  filter(Parameter %in% stats_to_return)
```

### View Time Tree
```{r fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
mcc_tree <- read.beast(here("local/run_2025-12-18/mcc.tree"))
tree_data <- as_tibble(mcc_tree)

p <- ggtree(mcc_tree, aes(color = posterior), mrsd="2023-11-08") +
  theme_tree2() +
  geom_tiplab(size = 3) +
  geom_text2(
    aes(label = ifelse(!isTip & !is.na(posterior), sprintf("%.2f", posterior), "")),
    hjust = -0.2, size = 2.8, color = "black"
  ) +
  scale_color_gradient(low = "gray60", high = "red") +
  labs(title = "VSV 2023 (MCC Tree)", color = "Posterior\nsupport") +
  theme(
    plot.margin = unit(c(0.5, 0.25, 0.5, 0.25), "cm"),
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14, face = "bold"),
    legend.position.inside = TRUE,
    legend.position = c(0.2, 0.7),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(size = 16, face = "bold")
  ) +
  scale_x_continuous(
    name = "  ",
    expand = c(0.01, 0)
  )

p

# ggsave(here("images/vsv_tree.pdf"), p, width = 16, height = 10, units = "in")
```

### GeoCoded Tree
```{r fig.width=10, fig.height=10, message=FALSE, warning=FALSE}
tree_loc <- read.beast(here("local/run_2025-12-18/location_mcc.tree"))

tip_meta <- as_tibble(tree_loc) %>% 
  select(label, location) %>%
  distinct()

loc_levels <- sort(unique(tip_meta$location))
loc_cols <- setNames(
  pals::trubetskoy(length(loc_levels)),
  loc_levels
)

loc_label_map <- c(
  SanBernardino  = "San Bernardino",
  SanDiego       = "San Diego",
  SanLuisObispo  = "San Luis Obispo",
  SantaBarbara   = "Santa Barbara"
)

tree_p <- ggtree(mcc_tree, mrsd = "2023-11-08") %<+% tip_meta +
  theme_tree2() +
  ggnewscale::new_scale_color() +
  geom_tippoint(
    aes(color = location),
    size = 4,
    alpha = 0.9
  ) +
  scale_color_manual(
  values = loc_cols,
  name   = "Location",
  labels = function(x) ifelse(x %in% names(loc_label_map),
                              loc_label_map[x],
                              x)
  ) +
  geom_tiplab(size = 3) +
  labs(title = " ") +
  theme(
    plot.margin = unit(c(0.15, 2.25, 0.15, 2.25), "cm"),
    legend.key.width = unit(1, "line"),
    legend.key.height = unit(2, "line"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 12, face = "bold"),
    legend.position.inside = TRUE,
    legend.position = c(0.08, 0.65),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold")
  ) +
  scale_x_continuous(
    name = "  ",
    expand = c(0.01, 0)
  )

tree_p

# ggsave(here("images/vsv_tree.pdf"), p, width = 16, height = 10, units = "in")
```


## Geographic Signal
Test for significance of time in the phylogeny
```{r}
# calculate genetic distances
gen_dist <- dist.dna(selected_seqs, model = "raw")

# geographic distances
geo_dist <- geosphere::distm(
  as.matrix(
    vsv_meta[, c("x", "y")]
    )
  )

# matrix names
rownames(geo_dist) <- vsv_meta$tip
colnames(geo_dist) <- vsv_meta$tip

# order
tips_order <- vsv_meta$tip
gen_dist <- as.matrix(gen_dist)[tips_order, tips_order]
geo_dist <- geo_dist[tips_order, tips_order]

```

### Mantel Test
Strong geographic signal
```{r}
mantel_result <- vegan::mantel(as.dist(gen_dist), 
                               as.dist(geo_dist), 
                               method = "pearson", 
                               permutations = 9999)
mantel_result
```

Save locations file for BEAST
```{r eval=FALSE}
locs_file <- vsv_meta %>%
  select(tip, county)

write.table(locs_file,
            file = here("local/vsv_locs.txt"),
            sep = "\t",
            row.names = FALSE,
            col.names = TRUE,
            quote = FALSE)
```

## Phylodynamics (Ne)
Estimating effective population size 
```{r message=FALSE, warning=FALSE, fig.width=15, fig.height=15}
phy <- as.phylo(mcc_tree)

phylodynamic_process(tree=phy,
                     mrsd = as.Date("2023-11-08"),
                     x_min = as.Date("2023-01-01"),
                     x_max = as.Date("2023-11-20")
                     )
```


## Phylogeography

### Transitions Network
Version with background
```{r message=FALSE, warning=FALSE, fig.height=15, fig.width=15}
plot_migration_network_terra(tree_loc,
                             vsv_meta,
                             ca_counties,
                             epi_start,
                             epi_end,
                             loc_cols)
```

version with simple background
```{r message=FALSE, warning=FALSE, fig.height=15, fig.width=15}
plot_migration_network(tree_loc,
                       vsv_meta,
                       ca_counties,
                       epi_start,
                       epi_end)
```

## Estimate Prevalence
Approximate prevalaence from case detections assuming an infectious period as described by [Humphreys, et. al 2024](https://www.mdpi.com/1999-4915/16/8/1315).

Credible interval from above paper implied vertebrate hosts remained infectious for 7.69–11.11 days post-infection.  
```{r}
CI_0.025_infect_period <- 7.69
CI_0.975_infect_period <- 11.11

# death rate
D_days <- (CI_0.025_infect_period+CI_0.975_infect_period)/2 # average
D_weeks <- D_days/7

df_prev <- comb_cases %>% 
  as.data.frame() %>%
  dplyr::select(-COUNTRY) %>%
  arrange(date)

# daily incidence
df_full <- df_prev %>%
  complete(date = seq(min(date), max(date), by = "day")) %>%
  mutate(inc = replace_na(inc, 0))

# infectiousness kernel
# kernel <- c(0.5, 0.5, 0) # weekly
kernel <- rep(1/2, D_days) # daily

N_pop <- 4e5 # approx horses in CA

df_full <- df_full %>%
  mutate(
    I_est = zoo::rollapply(
      inc,
      width = length(kernel),
      FUN = function(x) {
        k <- rev(kernel)[seq_along(x)]
        sum(k * x)
      },
      align = "right",
      fill = NA_real_,
      partial = TRUE
    ),
    prev_pop = I_est / N_pop# decimal rates for horse pop
  )

# whole count version
train_mod <- df_full %>%
  arrange(date) %>%
  mutate(time = as.integer(date - min(date)),
         prev = ceiling(if_else(is.na(I_est), 1, I_est))
  ) %>%
  select(time, prev)
```


### Scale Phylogeny
From units years to units days
```{r}
phy$edge.length <- phy$edge.length * 365.25 # to units days
```

### Outbreak Duration
```{r}
outbreak_dur <- max(comb_cases$date) - min(comb_cases$date) + 1
outbreak_dur

stop_time <- as.numeric(outbreak_dur) # units of days

# extract and align genomic and case data
align_genetic_data <- genetic_data(phy, stop_time, ptree_lag = epi_phy_lag)
align_genetic_data$n_coal[is.na(align_genetic_data$n_coal)] <- 0

death_rate <- 1/D_days # end of infectiousness takes D_days units of time
0.35/death_rate # basic R0

# Birth rate function - constant 0.3
constant <- function(t) {
  return(0.35)
}

```

## VSV Simulation  
The *pmmh()* function is provided by the [Gill et al 2025](https://academic.oup.com/jrsssc/advance-article/doi/10.1093/jrsssc/qlaf065/8362913) and the **EpiSky** R-package.  The version *pmmh2()* used here has been modified slightly to address indexing issues and unintentional conversion of dataframe objects to tibble class.
```{r eval=FALSE}
set.seed(1976)
vsv_simulation <- pmmh2(
  iter = 10000, # 10000 iterations,
  max_time = 300*60, # model run time (seconds)
  target_acceptance = 0.1, # sample acceptance rate
  sigma0 = 0.1,# initial smoothness parameter
  reporting_prob0 = 0.1, # probability of being detected/reported
  x0 = 3, # initial index infections
  genetic_data= align_genetic_data, # aligned tree and case reports
  death_rate = death_rate, # death rate as estimated above
  ptree = phy, # phylogeny
  ptree_lag = epi_phy_lag, # lag time between genomic and field reports
  sample_prevalence = train_mod, # case samples
  sigma_mean = 0.1, # exponential smoothness
  pobs_prior = "beta", # distribution
  pobs_alpha = 2, # beta param
  pobs_beta = 5, # beta param
  x0_prior = "nbinom", # prevalence distribution, negative binomial
  x0_mean = 5, # nbinom param
  x0_var = 10, # nbinom param
  n_particles = 1000, # particles
  ess_threshold = 500, # Effective Sample Size minimum
  resampling_scheme = "systematic", # resampling strategy
  backward_sim = TRUE # sample simulations
)

# save result
saveRDS(chain, here("local/est_R0/vsv_simulation_2025-12-30.rds"))
```

### Effective Reproduction and Prevalence
Read results
```{r}
vsv_simulation <- readRDS(here("local/est_R0/vsv_simulation_2025-12-30.rds"))
```

```{r fig.width=15, fig.height=15, message=FALSE, warning=FALSE}
plot_rt_prev(vsv_simulation, death_rate, 
             epi_start, epi_end,
             burn_in=2000)
```

